---
title: "Report 00"
author: "Nathan Shea Ouedraogo"
output: 
  html_document:
    toc: TRUE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
library(datasets)
library(tidyverse)
# load iris dataset
data(iris)
```

1. What are the column names and data types of the different columns in iris?

```{r}
# get column names
col_names <- colnames(iris)
print(col_names)

# map each column data type (class) to a list
data_types <- iris %>% map(class)
print(data_types)
```

2. How many rows and columns does iris have?

```{r}
print(nrow(iris))
print(ncol(iris))
```

3. Create a single vector (a new object) called ‘width’ that is the Sepal.Width column of iris.

```{r}
width <- iris$Sepal.Width
print(width)
```

4. What is the 100th value in your ‘width’ vector?

```{r}
print(width[100])
```

5. What is the last value in your ‘width’ vector? Can you write code that returns this value even if you don’t know how long ‘width’ is? 

```{r}
print(width[length(width)])
```

6. Select rows 10 to 20, with all columns in the iris dataset.

```{r}
iris_rows_10_to_20 <- iris %>% slice(10:20)
print(iris_rows_10_to_20)
```

7. Select rows 10 to 20 with only the Species, Petal.Width and Petal.Length. Can you do this two different ways?

```{r}
# method 1; just select rows
method_1 <- iris_rows_10_to_20 %>% select(Species, Petal.Width, Petal.Length)
print(method_1)

# method 2; "pull" out vectors and concatenate together
method_2 <- data.frame(
	Species      =	iris_rows_10_to_20 %>% pull(Species), 
	Petal.Width  = 	iris_rows_10_to_20 %>% pull(Petal.Width),
	Petal.Length =  iris_rows_10_to_20 %>% pull(Petal.Length)
)
print(method_2)
```

8. Select rows 1 to 10, 20, and 100 in the iris dataset.

```{r}
iris_row_slice <- iris %>% slice(1:10,20,100)
print(iris_row_slice)
```

9.  Select the first value in the Sepal.Length column of the iris dataset. Bonus - can you do this three different ways? 

```{r}
method_1 <- iris_row_slice$Sepal.Length[1]
print(method_1)
method_2 <- (iris_row_slice %>% pull(Sepal.Length))[1]
print(method_2)
method_3 <- first(iris_row_slice$Sepal.Length)
print(method_3)
```

10. Without running the following code in R, try to determine which of the following will return the first three rows of the Sepal.Length column in the iris data.frame? For each of the answers that do not work, see if you explain why!
   - a. iris[c(1, 2, 3), ‘Sepal.Length’]
   - b. iris[1, 2, 3, ‘Sepal.Length’]
   - c. iris[(1, 2, 3), ‘Sepal.Length’]
   - d. iris[‘Sepal.Length’, c(1, 2, 3)]

option a would work, since we concatenate 1,2,3 into a vector [1,2,3]
which represents row 1,2,3 of the Sepal.Length column.
